---
title: "P8105 HW2 Kad2243"
output: github_document
Date: 10/07/2026
---

# Problem 1
The purpose of this problem is to describe NYC subway data.

Load the tidyverse library: 
```{r setup,  message=FALSE }
library(tidyverse)
```

Read in the NYC Transit Subway Entrance and Exit data set:

```{r Transit Data import}
transit_df = read.csv("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
```

Looking at the data set:

```{r results = 'hide'}
head(transit_df)

tail(transit_df)
```

Changing variable names to snake case and selecting variables of interest:

Keep: line, station, name, station latitude / longitude, 
routes served, entry, vending, entrance type, and ADA compliance

Change Variable "entry" to logical var

```{r results = 'hide', message=FALSE, warning=FALSE}
transit_df = 
    janitor::clean_names(transit_df) %>%
    select(line, station_name, starts_with("route"), entry, vending, entrance_type, ada) %>%
    mutate( entry = case_match( entry, "YES" ~ TRUE, "NO" ~  FALSE))

```
