---
title: "P8105 HW2 Kad2243"
output: github_document
Date: 10/07/2026
---

# Problem 1
The purpose of this problem is to describe NYC subway data.

Load the tidyverse library: 
```{r setup,  message=FALSE }
library(tidyverse)
```

## Data import

Read in the NYC Transit Subway Entrance and Exit data set:

```{r Transit Data import}
transit_df = read.csv("./Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
```

Looking at the data set:

```{r results = 'hide'}
head(transit_df)

tail(transit_df)
```

## Data clean

Changing variable names to snake case and selecting variables of interest:

Keep: line, station, name, station latitude / longitude, 
routes served, entry, vending, entrance type, and ADA compliance

Change Variable "entry" to logical var

```{r results = 'hide', message=FALSE, warning=FALSE}
transit_df = 
    janitor::clean_names(transit_df) %>%
    select(line, station_name, starts_with("route"), entry, vending, entrance_type, ada) %>%
    mutate( entry = case_match( entry, "YES" ~ TRUE, "NO" ~  FALSE))

```

## Description of the NYC transit data set

The NYC Transit Subway Entrance and Exit Data set contains **`r nrow(transit_df)`** rows and **`r ncol(transit_df)`** columns.
The variables included in the set are:  **`r names(transit_df)`**. Line describes which major street or avenue is described, station name is the station or stop, route is the trains stopping at the station, etc. Following import of the raw data, variable names were renamed to include underscores, variables of interest were selected, and the variable "entry" was changed from YES/NO to TRUE/FALSE. The data are not particularly tidy, routes are captured across columns rather than as distinct observation, so route and route name are combined, with some route variables containing blanks for missing values and some containing NA. 

## Further description of data

- How many distinct stations are there? 

    There are **`r nrow(distinct(transit_df, line, station_name))`** distinct stations included in the data. 

- How many stations are ADA compliant?

    **`r nrow(filter(transit_df, ada == TRUE))`** stations are ADA compliant

- What proportion of station entrances / exits without vending allow entrance?

    **`r round(
              nrow(filter(transit_df, vending == "NO" & entry == TRUE)) / 
              nrow(filter(transit_df, vending == "NO")) ,
              2)
              `** of station entrances/exits without vending allow entrance 
            
## Distinct Route Number and Names

Reformat data so that route number and route name are distinct variables.

```{r}
transit_tidy = transit_df %>%
  mutate(across(route1:route11, as.character)) %>%
  pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  )
```

How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?

  - **`r nrow(distinct(filter(transit_tidy, route_name == "A"), line, station_name))`** stations serve the A train
  - **`r nrow(distinct(filter(transit_tidy, route_name == "A" & ada == TRUE), line, station_name))`** 
  stations serving the A train are ADA compliant

# Problem 2

The purpose of this problem is to load, clean, and merge multiple excel sheets to one tidy data set

Load the tidyverse library: 
```{r setup 2,  message=FALSE }
library(readxl)
```


## Data import

Read in all Trash Wheel data set:

```{r Trash Data import}
trash_mr = read_excel("./Data/202610 Trash Wheel Collection Data.xlsx", 
                      sheet = "Mr. Trash Wheel", 
                      range = cell_cols("A:N"))

trash_prof = read_excel("./Data/202610 Trash Wheel Collection Data.xlsx", 
                        sheet = "Professor Trash Wheel", 
                        range = cell_cols("A:M"))

trash_gwn = read_excel("./Data/202610 Trash Wheel Collection Data.xlsx", 
                       sheet = "Gwynnda the Good Wheel of the W", 
                       range = cell_cols("A:L"))
```

Looking at the data sets:

```{r results = 'hide'}
head(trash_mr)
tail(trash_mr)

head(trash_prof)
tail(trash_prof)

head(trash_gwn)
tail(trash_gwn)

```

## Data clean

For all Trash Wheel Data 
- Renaming variables using clean_names and manually renaming long variables
- omitting rows that do not include dumpster-specific data
- rounding sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
-Adding new var to each set for trash wheel identification
-Changing year to numeric, changing plastic bottles to numeric

```{r warning=FALSE}
trash_mr = 
    janitor::clean_names(trash_mr) %>%
    rename(weight_t = weight_tons, volume_y3 = volume_cubic_yards, cig = cigarette_butts ) %>%
    filter( dumpster != "") %>%
    mutate( sports_balls = as.integer(round(sports_balls,0)), trashwheel = "mr", year = as.numeric(year),)

trash_prof = 
    janitor::clean_names(trash_prof) %>%
    rename(weight_t = weight_tons, volume_y3 = volume_cubic_yards, cig = cigarette_butts ) %>%
    filter( dumpster != "") %>%
    mutate(trashwheel = "prof", plastic_bottles = as.numeric(plastic_bottles))

trash_gwn = 
    janitor::clean_names(trash_gwn) %>%
    rename(weight_t = weight_tons, volume_y3 = volume_cubic_yards, cig = cigarette_butts ) %>%
    filter( dumpster != "")%>%
    mutate(trashwheel = "gwn")
```


Creating a new merged dataset with the Trash Wheel sets

```{r}
trash_tidy=
  bind_rows(trash_mr, trash_prof, trash_gwn)
```

# Describing the data set 
- Note the number of observations in the resulting dataset, and give examples of key variables. 

  There are **`r nrow(trash_tidy)`** observations in the merged dataset, which includes data from Mr. Trash Wheel, Professor       Trash Wheel, and Gwynnda the Good Wheel of the W. Variables included in the dataset are: **`r names(trash_tidy)`**. Weight in   tons and cubic yards are key variables because they can quantify amounts of trash collected by each wheel. 

- For available data, what was the total weight of trash collected by Professor Trash Wheel? 
  
  The total weight of trash collected by Professor Trash Wheel was
  **`r sum(filter(trash_tidy, trashwheel == "prof")$weight_t)`**

- What was the total number of cigarette butts collected by Gwynnda in June of 2022?

  The total number of cigarette butts collected by Gwynnda in June of 2022
  **`r sum(filter(trash_tidy, trashwheel == "gwn"& month == "June" & year == 2022 )$cig)`**
  
  
# Problem 3

## Data import

Read in baseline and amyloid MCI sheets

```{r MCI import}
mci_bl = read.csv("./Data/data_mci/MCI_baseline.csv", skip =1)
mci_amy = read.csv("./Data/data_mci/mci_amyloid.csv", skip =1)
```


Looking at the data set:

```{r results = 'hide'}
head(mci_bl)
head(mci_amy)
```


## Data clean for baseline data

Changing variable names to snake case, changine male and female to character variables, changing apoe4 status to a logical variable, removing any participants with an age of MCI onset (did not meet inclusion criteria - had MCI at baseline)
- Filtering those that had MCI at baseline by taking age at onset and comparing to current age at baseline

```{r results = 'hide', message=FALSE, warning=FALSE}

mci_bl = 
    janitor::clean_names(mci_bl) %>%
    mutate(sex = case_match( sex, 0 ~ "female", 1 ~ "male"),
           apoe4 = case_match(apoe4, 0 ~ FALSE, 1 ~ TRUE),
           age_at_onset = as.numeric(age_at_onset)) %>%
    filter(!(age_at_onset <= current_age & !is.na(age_at_onset)))

```

# Data Description 

- How many participants were recruited, and of these how many develop MCI? 

    There were **`r nrow(mci_bl)`** participants recruited, with **`r nrow(filter(mci_bl, !is.na(age_at_onset)))`** who will        develop MCI

- What is the average baseline age? 

  The average baseline age is **`r round(mean(mci_bl$current_age), 2)`**

- What proportion of women in the study are APOE4 carriers?

  **`r nrow(filter(mci_bl, sex=="female" & apoe4 == TRUE))/nrow(filter(mci_bl, sex=="female"))`** of women in the study are APOE4   carriers. 
  
  

## Data clean for Biomarkers

Renaming variables with snake case, followed by pivot to a tidy data format.
Using times as variables, renaming "Baseline" to time 0 for consistency with
other time points. 
Changing variable types to numeric. 

```{r results = 'hide', message=FALSE, warning=FALSE}

mci_amy_long =
  janitor::clean_names(mci_amy) %>%
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    names_prefix = "time_",
    values_to = "amy_ratio") %>%
  mutate(
    time = as.numeric(recode(time, "baseline" = "0")),
    amy_ratio = as.numeric(amy_ratio),
    id = study_id)

```

- There are **`r nrow(distinct(mci_amy_long, study_id))`** distinct patients included in the biomarker data. 

- The mean amyloid ratio of participants included in the data set at time 0 is **`r round(mean(filter(mci_amy_long, time == 0)$amy_ratio, na.rm = TRUE), 3)`**, the mean at time 2 is  **`r round(mean(filter(mci_amy_long, time == 2)$amy_ratio, na.rm = TRUE), 3)`**, the mean at time 4 is  **`r round(mean(filter(mci_amy_long, time == 4)$amy_ratio, na.rm = TRUE), 3)`**, the mean at time 6 is  **`r round(mean(filter(mci_amy_long, time == 6)$amy_ratio, na.rm = TRUE), 3)`**, and the mean at time 8 is  **`r round(mean(filter(mci_amy_long, time == 8)$amy_ratio, na.rm = TRUE), 3)`**.

## Merge Data

Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings.

```{r results = 'hide'}
table(mci_amy_long$id)
table(mci_bl)
```

- Participants 484-495 appear in the amyloid biomarker dataset and not in the baseline dataset

Merging baseline and biomarker data sets by study ID
```{r}
mci_merge =
  left_join(mci_bl,mci_amy_long, by = "id")
```

  There are **`r nrow(mci_merge)`** observations in the merged dataset, which includes data from baseline and the biomarker set.There are **`r ncol(mci_merge)`** variables.  Variables included in the dataset are: **`r names(mci_merge)`**.


